Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-10-07 content.tex
Slashes -> and/or. Better discussion of call/cc.

Programs in higher-order languages heavily use function calls and method dispatch as part of their control flow.
%
Until recently, flow analyses for higher-order languages could not handle return flow precisely \citep{ianjohnson:vardoulakis-lmcs11, dvanhorn:Earl2010Pushdown}, which leads to several spurious paths (and thus false positives) due to the pervasiveness of function/method calls and subsequent returns.
%
These works, called CFA2 and PDCFA respectively, use pushdown automata as their approximation's target model of computation.
%
They are hence called ``pushdown analyses.''
%
\footnote{We will refer to the classic finite model analyses as ``regular analyses'' after regular languages.}
%
CFA2 and PDCFA have difficult details to easily apply to an off-the-shelf semantics --- especially if they feature non-local control transfer that breaks the pushdown model.
%%

@@

%%
A downside of the work providing \rackett{call/cc} is that it was an algorithmic change to the already complicated CFA2 --- there was no recipe for how to do this for one's favorite control operator.
%
This paper seeks to do just that with an operational view of what summarization is, in essence.
%
In other words, we give a concrete semantics to the tricks that the analyses in the literature use in the abstract, and maintain the original meaning of the language.
%
We also give intuitive analogies to well-established ideas/techniques so that the working semanticist can write a pushdown analysis for her language.
%
In order to demonstrate the applicability of this viewpoint, we show a new analysis for a language with composable control.
%
All of the semantics modeled in this paper are implemented in full detail in PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available online\footnote{\url{http://github.com/ianj/concrete-summaries}}.

@@

To turn this semantics into PDCFA's algorithm for constructing a Dyck state graph, given the appropriate $\alloc$, we apply a widening operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these components), in what is then called a $\System$.
%
We show only the memo and continuation table widening, since store widening would require rewriting newly-created contexts to use the global store following the step.
%
This rewriting is unnecessary with the store-counting per-wide-step technique presented in \citet{ianjohnson:oaam:icfp2013}, or the less precise timestamp approximation from \citet{dvanhorn:Shivers:1991:CFA}.
%
Since these components are shared, they are always the least upper bound of any respective component that the semantics produces (the intermediate set $I$) in order to stay sound.
%
We use a meta-function $\wn$, ``wide to narrow'', to shift between the two representations of states.

@@

Finally, we can show that even if the tables are shared amongst all states, they are precise enough to not interfere with the results.
\begin{theorem}[Equivalence with global tables]\label{thm:global}
  Tabular semantics has a bisimulation with memo/k-table-widened semantics.
\end{theorem}

@@

The result of this viewpoint is a sound, precise, and computable semantics for a ``pushdown approach'' to analyzing delimited, composable control.
%
To handle non-composable control, we can simply remove the prompts and keep $\approximate$.

@@

We prove this with the invariant that the continuation and meta-continuation have unique, well-founded unrollings in addition to the table invariants from \autoref{sec:pdcfa}, since entries in $\mktab$ are always fresh.

@@

The immediately related work is that of PDCFA \citep{dvanhorn:Earl2010Pushdown}, CFA2~\citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:Vardoulakis2011Pushdown}, and AAM ~\citep{dvanhorn:VanHorn2010Abstracting}, the first two of which we recreated in full detail.
%
The version of CFA2 that handles \rackett{call/cc} does not handle composable control, and is dependent on a restricted CPS representation.
%
They also had no way to tune the precision of their first class continuation approximation.
%
Since \rackett{call/cc} can be simulated with shift/reset\footnote{\RktPn{(}\RktSym{define}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{call/cc}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{f}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{shift}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{k}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{k}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{f}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{$\lambda$}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{x}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{shift}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{c}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{k}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{x}\RktPn{)}\RktPn{)}\RktPn{)}\RktPn{)}\RktPn{)}\RktPn{)}\RktPn{)}}, this work supercedes theirs (the space complexity difference imposed by the simulation is non-issue in the abstract).
%
The extended version of PDCFA that does garbage collection~\citep{dvanhorn:Earl2012Introspective} also fits into our model;
the addresses that the stack keeps alive can be accumulated by ``reading through'' the continuation table, building up the set of addresses in each portion of the stack that we come across.
%%

%%
Pushdown models have existed in the first-order static analysis literature ~\citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}\citep{ianjohnson:reps:pushdown:1995}, and the first-order model checking literature \citep{ianjohnson:bouajiani:esparza:pushdown:1997}, for some time.
%
The higher-order setting imposes additional challenges that make their methods difficult to adapt.
%
The most important constraint is that we can't know all call-sites of a function/method before the analysis begins, which their methods heavily rely on.
%%

>>>>

Programs in higher-order languages heavily use function calls and method dispatch as part of their control flow.
%
Until recently, flow analyses for higher-order languages could not handle return flow precisely \citep{ianjohnson:vardoulakis-lmcs11, dvanhorn:Earl2010Pushdown}, which leads to several spurious paths (and thus false positives) due to the pervasiveness of function or method calls and subsequent returns.
%
These works, called CFA2 and PDCFA respectively, use pushdown automata as their approximation's target model of computation.
%
They are hence called ``pushdown analyses.''
%
\footnote{We will refer to the classic finite model analyses as ``regular analyses'' after regular languages.}
%
CFA2 and PDCFA have difficult details to easily apply to an off-the-shelf semantics --- especially if they feature non-local control transfer that breaks the pushdown model.
%%

@@

%%
A downside of the work providing \rackett{call/cc} is that it was an algorithmic change to the already complicated CFA2 --- there was no recipe for how to do this for one's favorite control operator.
%
This paper seeks to do just that with an operational view of what summarization is, in essence.
%
In other words, we give a concrete semantics to the tricks that the analyses in the literature use in the abstract, and maintain the original meaning of the language.
%
We also give intuitive analogies to well-established ideas and techniques so that the working semanticist can write a pushdown analysis for her language.
%
In order to demonstrate the applicability of this viewpoint, we show a new analysis for a language with composable control.
%
All of the semantics modeled in this paper are implemented in full detail in PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available online\footnote{\url{http://github.com/ianj/concrete-summaries}}.

@@

To turn this semantics into PDCFA's algorithm for constructing a Dyck state graph, given the appropriate $\alloc$, we apply a widening operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these components), in what is then called a $\System$. We use a meta-function $\wn$, ``wide to narrow,'' to shift between the two representations of states.
%%

%%
We show only the memo and continuation table widening, since store widening would require rewriting newly-created contexts to use the global store following the step.
%
This rewriting is unnecessary with the store-counting per-wide-step technique presented in \citet{ianjohnson:oaam:icfp2013}, or the less precise timestamp approximation from \citet{dvanhorn:Shivers:1991:CFA}.
%
Since these components are shared, they are always the least upper bound of any respective component that the semantics produces (the intermediate set $I$) in order to stay sound.

@@

Finally, we can show that even if the tables are shared amongst all states, they are precise enough to not interfere with the results.
\begin{theorem}[Equivalence with global tables]\label{thm:global}
  Tabular semantics has a bisimulation with memo-and-k-table-widened semantics.
\end{theorem}

@@

The result of this viewpoint is a sound, precise, and computable semantics for a ``pushdown approach'' to analyzing delimited, composable control.

@@

We prove this with the invariant that the continuation and meta-continuation have unique unrollings in addition to the table invariants from \autoref{sec:pdcfa}, since entries in $\mktab$ are always fresh.

@@

The immediately related work is that of PDCFA \citep{dvanhorn:Earl2010Pushdown}, CFA2~\citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:Vardoulakis2011Pushdown}, and AAM ~\citep{dvanhorn:VanHorn2010Abstracting}, the first two of which we recreated in full detail.
%
The version of CFA2 that handles \rackett{call/cc} does not handle composable control, and is dependent on a restricted CPS representation.
%
They also had no way to tune the precision of their first class continuation approximation.
%
Our semantics adapts to \rackett{call/cc} by removing the meta-continuation operations, and thus this work supercedes theirs.
%
The extended version of PDCFA that does garbage collection~\citep{dvanhorn:Earl2012Introspective} also fits into our model;
the addresses that the stack keeps alive can be accumulated by ``reading through'' the continuation table, building up the set of addresses in each portion of the stack that we come across.
%%

%%
Pushdown models have existed in the first-order static analysis literature ~\citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}\citep{ianjohnson:reps:pushdown:1995}, and the first-order model checking literature \citep{ianjohnson:bouajiani:esparza:pushdown:1997}, for some time.
%
The higher-order setting imposes additional challenges that make their methods difficult to adapt.
%
The most important constraint is that we can't know all call-sites of a function or method before the analysis begins, which their methods heavily rely on.
%%
